### 2018-02-01 submission/manuscript.Rmd
intro modifications

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

@@

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that differences, based on DNA polymerase, in the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

>>>>

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

@@

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline regardless of DNA polymerase used [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

We amplified the V4 region of the 16S rRNA gene in both fecal and mock community samples using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA polymerase. First, we tested if we could identify differences in the number of OTUs between the different HiFi DNA polymerase for both the fecal and mock samples. Next, since there were differences based on HiFi DNA polymerase, we examined if a community measure such as the Bray-Curtis index would be affected. After assessing these community measures we then examined if the different HiFi DNA polymerase had different per base error rates as well as chimera prevalence. Finally, since differences in chimera prevalence, based on HiFi DNA polymerase, could not be completely removed using bioinformatic approaches we tested whether the chimera prevalence correlated well with total number of observed OTUs.

### 2018-02-01 submission/manuscript.tex
intro modifications

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced bias, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include, but are not limited to, preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. Thus, for these overlapping considerations biases
identified for 16S rRNA gene sequencing will also likely influence
metagenomic sequencing results.

@@

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that differences, based on DNA polymerase, in the number of
OTUs generated were not easily removed using the authors chosen
bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families (e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples
using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
polymerase. First, we tested if we could identify differences in the
number of OTUs between the different HiFi DNA polymerase for both the
fecal and mock samples. Next, since there were differences based on HiFi
DNA polymerase, we examined if a community measure such as the
Bray-Curtis index would be affected. After assessing these community
measures we then examined if the different HiFi DNA polymerase had
different per base error rates as well as chimera prevalence. Finally,
since differences in chimera prevalence, based on HiFi DNA polymerase,
could not be completely removed using bioinformatic approaches we tested
whether the chimera prevalence correlated well with total number of
observed OTUs.

>>>>

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced biases, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include preservation media, storage conditions, DNA
extraction kit, PCR, and sequence library preparation. Thus, for these
overlapping considerations biases identified for 16S rRNA gene
sequencing will also likely influence metagenomic sequencing results.

@@

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that the number of OTUs generated were not easily removed
using the authors chosen bioinformatic pipeline regardless of DNA
polymerase used (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families (e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the V4 region of the 16S rRNA gene in both fecal and mock
community samples using either Accuprime, Kappa, Q5, Phusion, or
Platinum HiFi DNA polymerase. First, we tested if we could identify
differences in the number of OTUs between the different HiFi DNA
polymerase for both the fecal and mock samples. Next, since there were
differences based on HiFi DNA polymerase, we examined if a community
measure such as the Bray-Curtis index would be affected. After assessing
these community measures we then examined if the different HiFi DNA
polymerase had different per base error rates as well as chimera
prevalence. Finally, since differences in chimera prevalence, based on
HiFi DNA polymerase, could not be completely removed using bioinformatic
approaches we tested whether the chimera prevalence correlated well with
total number of observed OTUs.
